cardinalR: Generating interesting high-dimensional data structures

### 2025-09-23 paper-cardinalR.Rmd
Removed all the tables

The main arguments of the `gen_multicluster()` function are shown in Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:main-tb-html)', '\\@ref(tab:main-tb-pdf)'))`.

```{r}
main_tb <- tibble(arg = c("n",
                          "p",
                          "k",
                          "loc",
                          "scale",
                          "shape",
                          "rotation",
                          "is_bkg"),
                  type = c("numeric (vector)",
                           "numeric",
                           "numeric",
                           "numeric (matrix)",
                           "numeric (vector)",
                           "character (vector)",
                           "numeric (list)",
                           "boolean"),
                        exp = c("Number of points in each cluster.",
                                "Number of dimensions.",
                                "Number of clusters.",
                                "Locations/centroids of clusters.",
                                "Scaling factors of clusters.",
                                "Shapes of clusters.",
                                "Plane and the corresponding angle along that plane for each cluster.",
                                "Background noise should exist or not."))
```

```{r main-tb-html, eval=knitr::is_html_output()}
main_tb |> 
  kable(caption = "The main arguments for `gen_multicluster()`.", col.names = c("Argument", "Type", "Explanation")) 
```

```{r main-tb-pdf, eval=knitr::is_latex_output()}
main_tb |> 
  kable(caption = "The main arguments for gen\\_multicluster().", format="latex", col.names = c("Argument", "Type", "Explanation"), booktabs = T, table.pos = "H")  |>
  column_spec(1, width = "4cm") |>
  column_spec(2, width = "4cm") |>
  column_spec(3, width = "8cm")
```

@@

A branching structure (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch)"} else { "Figure \\@ref(fig:branch-proj)"})`) captures trajectories that diverge or bifurcate from a common origin, similar to processes such as cell differentiation in biology (@trapnell2014). We introduce a set of data generation functions specifically designed to simulate high-dimensional branching structures with various geometries, numbers of points (`n`), and number of branches (`k`). Although these functions can generate multiple branches, they do not produce a formal *multicluster* dataset: the branches form a single connected structure, with multiple visually distinct arms rather than independent clusters. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:branching-tb-html)', '\\@ref(tab:branching-tb-pdf)'))` outlines these functions. The main arguments of the functions described in Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:arg-branching-tb-html)', '\\@ref(tab:arg-branching-tb-pdf)'))`.

```{r}
branch_tb <- tibble(fun = c("gen_expbranches",
                             "gen_linearbranches",
                             "gen_curvybranches", 
                             "gen_orglinearbranches", 
                             "gen_orgcurvybranches"), 
                      exp = c("Exponential shaped branches.",
                               "Linear shaped branches.", 
                               "Curvy shaped branches.",
                               "Linear shaped branches originated in one point.",
                               "Curvy shaped branches originated in one point."))
```

```{r branching-tb-html, eval=knitr::is_html_output()}
branch_tb |> 
  kable(caption = "cardinalR branching data generation functions", col.names = c("Function", "Explanation")) 
```

```{r branching-tb-pdf, eval=knitr::is_latex_output()}
branch_tb |> 
  kable(caption = "cardinalR branching data generation functions", format="latex", col.names = c("Function", "Explanation"), booktabs = T)  |>
  column_spec(1, width = "4cm") |>
  column_spec(2, width = "8cm")
```

```{r}
arg_branching_tb <- tibble(arg = c("n",
                          "p",
                          "k"), 
                        exp = c("A numeric value representing the number of points.",
                                "A numeric value representing the number of dimensions.",
                                "A numeric value representing the number of clusters."))
```

```{r arg-branching-tb-html, eval=knitr::is_html_output()}
arg_branching_tb |> 
  kable(caption = "The main arguments for branching shape generators.", col.names = c("Argument", "Explanation")) 
```

```{r arg-branching-tb-pdf, eval=knitr::is_latex_output()}
arg_branching_tb |> 
  kable(caption = "The main arguments for branching shape generators.", format="latex", col.names = c("Argument", "Explanation"), booktabs = T)  |>
  column_spec(1, width = "4cm") |>
  column_spec(2, width = "8cm")
```

@@

A cube structure (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:cube)"} else { "Figure \\@ref(fig:cube-proj)"})`) represents uniformly or systematically distributed points within a high-dimensional hypercube, providing a useful framework for assessing how well algorithms preserve uniformity, spacing, and boundary properties in high dimensions. We provide a set of functions to generate high-dimensional cube structures with flexible configurations, including regular grids, uniform random points, and cubes with missing regions or holes. These structures are valuable for testing the ability of algorithms to maintain uniform spacing or to detect gaps in the data. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:cube-tb-html)', '\\@ref(tab:cube-tb-pdf)'))` outlines these functions and their purposes.

@@

```{r}
cube_tb <- tibble(fun = c("gen_gridcube",
                          "gen_unifcube",
                          "gen_cubehole"), 
                      exp = c("Cube with specified grid points along each axes.",
                              "Cube with uniform points.", 
                              "Cube with a hole."))
```

```{r cube-tb-html, eval=knitr::is_html_output()}
cube_tb |> 
  kable(caption = "cardinalR cube data generation functions", col.names = c("Function", "Explanation")) 
```

```{r cube-tb-pdf, eval=knitr::is_latex_output()}
cube_tb |> 
  kable(caption = "cardinalR cube data generation functions", format="latex", col.names = c("Function", "Explanation"), booktabs = T)  |>
  column_spec(1, width = "4cm") |>
  column_spec(2, width = "8cm")
```

@@

A polynomial structure (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:poly)"} else { "Figure \\@ref(fig:poly-proj)"})`) generates data points that follow non-linear curvilinear relationships, such as quadratic or cubic trends, in high-dimensional space. These patterns are useful for evaluating how well algorithms capture smooth, non-linear trajectories and curvature in the data. We provide functions for generating quadratic and cubic structures, enabling controlled experiments with different degrees of polynomial complexity. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:polynomial-tb-html)', '\\@ref(tab:polynomial-tb-pdf)'))` summarizes these functions and their purposes.

```{r}
polynomial_tb <- tibble(fun = c("gen_quadratic",
                                "gen_cubic"), 
                        exp = c("Quadratic pattern.",
                                "Cubic pattern."))
```

```{r polynomial-tb-html, eval=knitr::is_html_output()}
polynomial_tb |> 
  kable(caption = "cardinalR polynomial data generation functions", col.names = c("Function", "Explanation")) 
```

```{r polynomial-tb-pdf, eval=knitr::is_latex_output()}
polynomial_tb |> 
  kable(caption = "cardinalR polynomial data generation functions", format="latex", col.names = c("Function", "Explanation"), booktabs = T)  |>
  column_spec(1, width = "4cm") |>
  column_spec(2, width = "8cm")
```

@@

A pyramid structure (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:pyr)"} else { "Figure \\@ref(fig:pyr-proj)"})`) represents data arranged around a central apex and base, useful for exploring how algorithms handle pointed or layered geometries in high-dimensional space. The functions provided allow users to generate pyramids with rectangular, triangular, and star-shaped bases, and sharp or blunted apexes. Additionally, it is possible to create a pyramid with a fractal-like internal structure, enabling the study of non-convex and sparse regions. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:pyramid-tb-html)', '\\@ref(tab:pyramid-tb-pdf)'))` summarizes these functions.

@@

```{r}
pyramid_tb <- tibble(fun = c("gen_pyrrect",
                             "gen_pyrtri",
                             "gen_pyrstar",
                             "gen_pyrfrac"), 
                        exp = c("Rectangular-base pyramid, with a sharp or blunted apex.",
          "Triangular-base pyramid, with a sharp or blunted apex.",
          "Star-shaped base pyramid, with a sharp or blunted apex.",
          "Pyramid containing triangular pyramid-shaped holes."))
```

```{r pyramid-tb-html, eval=knitr::is_html_output()}
pyramid_tb |> 
  kable(caption = "cardinalR pyramid data generation functions", col.names = c("Function", "Explanation")) 
```

```{r pyramid-tb-pdf, eval=knitr::is_latex_output()}
pyramid_tb |> 
  kable(caption = "cardinalR pyramid data generation functions", format="latex", col.names = c("Function", "Explanation"), booktabs = T)  |>
  column_spec(1, width = "4cm") |>
  column_spec(2, width = "8cm")
```

@@

An S-curve structure (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:scurve)"} else { "Figure \\@ref(fig:scurve-proj)"})`) simulates data that lies along a smooth, non-linear manifold. The functions generate both the standard S-curve shape and, a S-curve variant with structured hole that introduce missing or incomplete region. These variations are useful for evaluating how well algorithms capture non-linear geometry and handle incomplete manifolds in high-dimensional data. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:scurve-tb-html)', '\\@ref(tab:scurve-tb-pdf)'))` summarizes these functions.

```{r}
scurve_tb <- tibble(fun = c("gen_scurve",
                            "gen_scurvehole"), 
                        exp = c("S-curve.",
                                "S-curve with a hole."))
```

```{r scurve-tb-html, eval=knitr::is_html_output()}
scurve_tb |> 
  kable(caption = "cardinalR S-curve data generation functions", col.names = c("Function", "Explanation")) 
```

```{r scurve-tb-pdf, eval=knitr::is_latex_output()}
scurve_tb |> 
  kable(caption = "cardinalR S-curve data generation functions", format="latex", col.names = c("Function", "Explanation"), booktabs = T)  |>
  column_spec(1, width = "4cm") |>
  column_spec(2, width = "8cm")
```

@@

Sphere-shaped structures (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere)"} else { "Figure \\@ref(fig:sphere-proj)"})`) are useful for evaluating how dimension reduction and clustering algorithms handle curved, symmetric manifolds in high-dimensional spaces. The functions generate a variety of spherical forms, including simple circles, uniform spheres, grid-based spheres, and complex arrangements like clustered spheres within a larger sphere. The first few coordinates define the main geometric form (circle, cycle, sphere, or hemisphere), while higher-dimensional embeddings are achieved by adding noise dimensions. For the circle and curvy cycle, this occurs when $p > 2$ and $p > 3$, respectively, while for spheres and hemispheres it occurs when $p > 3$ or $p > 4$. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:sphere-tb-html)', '\\@ref(tab:sphere-tb-pdf)'))` summarizes these functions.

```{r}
sphere_tb <- tibble(fun = c("gen_circle",
                            "gen_curvycycle",
                            "gen_unifsphere",
                            "gen_gridedsphere",
                            "gen_clusteredspheres",
                            "gen_hemisphere"), 
                        exp = c("Circle.",
                                "Curvy cell cycle.",
                                "Uniform sphere.",
                                "Grided sphere.",
                                "Multiple small spheres within a big sphere.",
                                "Hemisphere."))
```

```{r sphere-tb-html, eval=knitr::is_html_output()}
sphere_tb |> 
  kable(caption = "cardinalR sphere data generation functions", col.names = c("Function", "Explanation")) 
```

```{r sphere-tb-pdf, eval=knitr::is_latex_output()}
sphere_tb |> 
  kable(caption = "cardinalR sphere data generation functions", format="latex", col.names = c("Function", "Explanation"), booktabs = T)  |>
  column_spec(1, width = "4cm") |>
  column_spec(2, width = "8cm")
```

@@

Trigonometric-based structures provide flexible ways to simulate complex curved patterns and spirals that often arise in real-world high-dimensional data, such as in biological trajectories, or physical systems (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:triginometric)"} else { "Figure \\@ref(fig:triginometric-proj)"})`). The first few coordinates define the main geometric structure (e.g., crescent, cylinder, spiral, or helix), while for higher-dimensional embeddings, additional noise dimensions are appended: for the crescent shape this occurs when $p > 2$ (adding $X_3$ through $X_p$), and for all other trigonometric shapes when $p > 4$ (adding $X_5$ through $X_p$). These structures are particularly valuable for testing how well dimension reduction and clustering algorithms preserve intricate geometric and topological features. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:trigonometric-tb-html)', '\\@ref(tab:trigonometric-tb-pdf)'))` summarizes these functions.

```{r}
trigonometric_tb <- tibble(fun = c("gen_crescent",
                            "gen_curvycylinder",
                            "gen_sphericalspiral",
                            "gen_helicalspiral",
                            "gen_conicspiral",
                            "gen_nonlinear"), 
                        exp = c("Crescent pattern.",
                                "Curvy cylinder.",
                                "Spherical spiral.",
                                "Helical spiral.",
                                "Conic spiral.",
                                "Nonlinear hyperbola."))
```

```{r trigonometric-tb-html, eval=knitr::is_html_output()}
trigonometric_tb |> 
  kable(caption = "cardinalR trigonometric data generation functions", col.names = c("Function", "Explanation")) 
```

```{r trigonometric-tb-pdf, eval=knitr::is_latex_output()}
trigonometric_tb |> 
  kable(caption = "cardinalR trigonometric data generation functions", format="latex", col.names = c("Function", "Explanation"), booktabs = T)  |>
  column_spec(1, width = "4cm") |>
  column_spec(2, width = "8cm")
```

@@

```{r}
noise_fun_tb <- tibble(fun = c("gen_noisedims",
                            "gen_wavydims1",
                            "gen_wavydims2",
                            "gen_wavydims3"), 
                        exp = c("Gaussian noise dimensions with optional mean and standard deviation.",
                                "Wavy noise dimensions based on a user-specified theta sequence with added jitter.",
                                "Wavy noise dimensions using polynomial transformations of an existing dimension vector.",
                                "Wavy noise dimensions using a combination of polynomial and sine transformations based on the first three dimensions of a dataset."))
```

```{r noise-tb-html, eval=knitr::is_html_output()}
noise_fun_tb |> 
  kable(caption = "cardinalR noise dimensions generation functions", col.names = c("Function", "Explanation")) 
```

```{r noise-tb-pdf, eval=knitr::is_latex_output()}
noise_fun_tb |> 
  kable(caption = "cardinalR noise dimensions generation functions", format="latex", col.names = c("Function", "Explanation"), booktabs = T)  |>
  column_spec(1, width = "4cm") |>
  column_spec(2, width = "8cm")
```

The `gen_noisedims(n, p, m, s)` function generates \$p\$ independent Gaussian noise dimensions,

>>>>

@@

A branching structure (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch)"} else { "Figure \\@ref(fig:branch-proj)"})`) captures trajectories that diverge or bifurcate from a common origin, similar to processes such as cell differentiation in biology (@trapnell2014). We introduce a set of data generation functions specifically designed to simulate high-dimensional branching structures with various geometries, numbers of points (`n`), and number of branches (`k`). Although these functions can generate multiple branches, they do not produce a formal *multicluster* dataset: the branches form a single connected structure, with multiple visually distinct arms rather than independent clusters.

@@

A cube structure (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:cube)"} else { "Figure \\@ref(fig:cube-proj)"})`) represents uniformly or systematically distributed points within a high-dimensional hypercube, providing a useful framework for assessing how well algorithms preserve uniformity, spacing, and boundary properties in high dimensions. We provide a set of functions to generate high-dimensional cube structures with flexible configurations, including regular grids, uniform random points, and cubes with missing regions or holes. These structures are valuable for testing the ability of algorithms to maintain uniform spacing or to detect gaps in the data.

@@

@@

A polynomial structure (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:poly)"} else { "Figure \\@ref(fig:poly-proj)"})`) generates data points that follow non-linear curvilinear relationships, such as quadratic or cubic trends, in high-dimensional space. These patterns are useful for evaluating how well algorithms capture smooth, non-linear trajectories and curvature in the data. We provide functions for generating quadratic and cubic structures, enabling controlled experiments with different degrees of polynomial complexity.

@@

A pyramid structure (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:pyr)"} else { "Figure \\@ref(fig:pyr-proj)"})`) represents data arranged around a central apex and base, useful for exploring how algorithms handle pointed or layered geometries in high-dimensional space. The functions provided allow users to generate pyramids with rectangular, triangular, and star-shaped bases, and sharp or blunted apexes. Additionally, it is possible to create a pyramid with a fractal-like internal structure, enabling the study of non-convex and sparse regions.

@@

@@

An S-curve structure (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:scurve)"} else { "Figure \\@ref(fig:scurve-proj)"})`) simulates data that lies along a smooth, non-linear manifold. The functions generate both the standard S-curve shape and, a S-curve variant with structured hole that introduce missing or incomplete region. These variations are useful for evaluating how well algorithms capture non-linear geometry and handle incomplete manifolds in high-dimensional data.

@@

Sphere-shaped structures (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere)"} else { "Figure \\@ref(fig:sphere-proj)"})`) are useful for evaluating how dimension reduction and clustering algorithms handle curved, symmetric manifolds in high-dimensional spaces. The functions generate a variety of spherical forms, including simple circles, uniform spheres, grid-based spheres, and complex arrangements like clustered spheres within a larger sphere. The first few coordinates define the main geometric form (circle, cycle, sphere, or hemisphere), while higher-dimensional embeddings are achieved by adding noise dimensions. For the circle and curvy cycle, this occurs when $p > 2$ and $p > 3$, respectively, while for spheres and hemispheres it occurs when $p > 3$ or $p > 4$.

@@

Trigonometric-based structures provide flexible ways to simulate complex curved patterns and spirals that often arise in real-world high-dimensional data, such as in biological trajectories, or physical systems (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:triginometric)"} else { "Figure \\@ref(fig:triginometric-proj)"})`). The first few coordinates define the main geometric structure (e.g., crescent, cylinder, spiral, or helix), while for higher-dimensional embeddings, additional noise dimensions are appended: for the crescent shape this occurs when $p > 2$ (adding $X_3$ through $X_p$), and for all other trigonometric shapes when $p > 4$ (adding $X_5$ through $X_p$). These structures are particularly valuable for testing how well dimension reduction and clustering algorithms preserve intricate geometric and topological features.

@@

The `gen_noisedims(n, p, m, s)` function generates $p$ independent Gaussian noise dimensions,

### 2025-09-23 paper-cardinalR.tex
Removed all the tables

The main arguments of the \texttt{gen\_multicluster()} function are shown in Table \ref{tab:main-tb-pdf}.

\caption{\label{tab:main-tb-pdf}The main arguments for gen\_multicluster().}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{8cm}}
\toprule
Argument & Type & Explanation\\
\midrule
n & numeric (vector) & Number of points in each cluster.\\
p & numeric & Number of dimensions.\\
k & numeric & Number of clusters.\\
loc & numeric (matrix) & Locations/centroids of clusters.\\
scale & numeric (vector) & Scaling factors of clusters.\\
\addlinespace
shape & character (vector) & Shapes of clusters.\\
rotation & numeric (list) & Plane and the corresponding angle along that plane for each cluster.\\
is\_bkg & boolean & Background noise should exist or not.\\
\bottomrule
\end{tabular}
\end{table}

@@

A branching structure (Figure \ref{fig:branch-proj}) captures trajectories that diverge or bifurcate from a common origin, similar to processes such as cell differentiation in biology (\citet{trapnell2014}). We introduce a set of data generation functions specifically designed to simulate high-dimensional branching structures with various geometries, numbers of points (\texttt{n}), and number of branches (\texttt{k}). Although these functions can generate multiple branches, they do not produce a formal \emph{multicluster} dataset: the branches form a single connected structure, with multiple visually distinct arms rather than independent clusters. Table \ref{tab:branching-tb-pdf} outlines these functions. The main arguments of the functions described in Table \ref{tab:arg-branching-tb-pdf}.

\caption{\label{tab:branching-tb-pdf}cardinalR branching data generation functions}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{8cm}}
\toprule
Function & Explanation\\
\midrule
gen\_expbranches & Exponential shaped branches.\\
gen\_linearbranches & Linear shaped branches.\\
gen\_curvybranches & Curvy shaped branches.\\
gen\_orglinearbranches & Linear shaped branches originated in one point.\\
gen\_orgcurvybranches & Curvy shaped branches originated in one point.\\
\bottomrule
\end{tabular}
\end{table}

\caption{\label{tab:arg-branching-tb-pdf}The main arguments for branching shape generators.}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{8cm}}
\toprule
Argument & Explanation\\
\midrule
n & A numeric value representing the number of points.\\
p & A numeric value representing the number of dimensions.\\
k & A numeric value representing the number of clusters.\\
\bottomrule
\end{tabular}
\end{table}

@@

A cube structure (Figure \ref{fig:cube-proj}) represents uniformly or systematically distributed points within a high-dimensional hypercube, providing a useful framework for assessing how well algorithms preserve uniformity, spacing, and boundary properties in high dimensions. We provide a set of functions to generate high-dimensional cube structures with flexible configurations, including regular grids, uniform random points, and cubes with missing regions or holes. These structures are valuable for testing the ability of algorithms to maintain uniform spacing or to detect gaps in the data. Table \ref{tab:cube-tb-pdf} outlines these functions and their purposes.

@@

\caption{\label{tab:cube-tb-pdf}cardinalR cube data generation functions}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{8cm}}
\toprule
Function & Explanation\\
\midrule
gen\_gridcube & Cube with specified grid points along each axes.\\
gen\_unifcube & Cube with uniform points.\\
gen\_cubehole & Cube with a hole.\\
\bottomrule
\end{tabular}
\end{table}

@@

A polynomial structure (Figure \ref{fig:poly-proj}) generates data points that follow non-linear curvilinear relationships, such as quadratic or cubic trends, in high-dimensional space. These patterns are useful for evaluating how well algorithms capture smooth, non-linear trajectories and curvature in the data. We provide functions for generating quadratic and cubic structures, enabling controlled experiments with different degrees of polynomial complexity. Table \ref{tab:polynomial-tb-pdf} summarizes these functions and their purposes.

\caption{\label{tab:polynomial-tb-pdf}cardinalR polynomial data generation functions}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{8cm}}
\toprule
Function & Explanation\\
\midrule
gen\_quadratic & Quadratic pattern.\\
gen\_cubic & Cubic pattern.\\
\bottomrule
\end{tabular}
\end{table}

@@

A pyramid structure (Figure \ref{fig:pyr-proj}) represents data arranged around a central apex and base, useful for exploring how algorithms handle pointed or layered geometries in high-dimensional space. The functions provided allow users to generate pyramids with rectangular, triangular, and star-shaped bases, and sharp or blunted apexes. Additionally, it is possible to create a pyramid with a fractal-like internal structure, enabling the study of non-convex and sparse regions. Table \ref{tab:pyramid-tb-pdf} summarizes these functions.

@@

\caption{\label{tab:pyramid-tb-pdf}cardinalR pyramid data generation functions}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{8cm}}
\toprule
Function & Explanation\\
\midrule
gen\_pyrrect & Rectangular-base pyramid, with a sharp or blunted apex.\\
gen\_pyrtri & Triangular-base pyramid, with a sharp or blunted apex.\\
gen\_pyrstar & Star-shaped base pyramid, with a sharp or blunted apex.\\
gen\_pyrfrac & Pyramid containing triangular pyramid-shaped holes.\\
\bottomrule
\end{tabular}
\end{table}

@@

An S-curve structure (Figure \ref{fig:scurve-proj}) simulates data that lies along a smooth, non-linear manifold. The functions generate both the standard S-curve shape and, a S-curve variant with structured hole that introduce missing or incomplete region. These variations are useful for evaluating how well algorithms capture non-linear geometry and handle incomplete manifolds in high-dimensional data. Table \ref{tab:scurve-tb-pdf} summarizes these functions.

\caption{\label{tab:scurve-tb-pdf}cardinalR S-curve data generation functions}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{8cm}}
\toprule
Function & Explanation\\
\midrule
gen\_scurve & S-curve.\\
gen\_scurvehole & S-curve with a hole.\\
\bottomrule
\end{tabular}
\end{table}

@@

Sphere-shaped structures (Figure \ref{fig:sphere-proj}) are useful for evaluating how dimension reduction and clustering algorithms handle curved, symmetric manifolds in high-dimensional spaces. The functions generate a variety of spherical forms, including simple circles, uniform spheres, grid-based spheres, and complex arrangements like clustered spheres within a larger sphere. The first few coordinates define the main geometric form (circle, cycle, sphere, or hemisphere), while higher-dimensional embeddings are achieved by adding noise dimensions. For the circle and curvy cycle, this occurs when \(p > 2\) and \(p > 3\), respectively, while for spheres and hemispheres it occurs when \(p > 3\) or \(p > 4\). Table \ref{tab:sphere-tb-pdf} summarizes these functions.

\caption{\label{tab:sphere-tb-pdf}cardinalR sphere data generation functions}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{8cm}}
\toprule
Function & Explanation\\
\midrule
gen\_circle & Circle.\\
gen\_curvycycle & Curvy cell cycle.\\
gen\_unifsphere & Uniform sphere.\\
gen\_gridedsphere & Grided sphere.\\
gen\_clusteredspheres & Multiple small spheres within a big sphere.\\
\addlinespace
gen\_hemisphere & Hemisphere.\\
\bottomrule
\end{tabular}
\end{table}

@@

Trigonometric-based structures provide flexible ways to simulate complex curved patterns and spirals that often arise in real-world high-dimensional data, such as in biological trajectories, or physical systems (Figure \ref{fig:triginometric-proj}). The first few coordinates define the main geometric structure (e.g., crescent, cylinder, spiral, or helix), while for higher-dimensional embeddings, additional noise dimensions are appended: for the crescent shape this occurs when \(p > 2\) (adding \(X_3\) through \(X_p\)), and for all other trigonometric shapes when \(p > 4\) (adding \(X_5\) through \(X_p\)). These structures are particularly valuable for testing how well dimension reduction and clustering algorithms preserve intricate geometric and topological features. Table \ref{tab:trigonometric-tb-pdf} summarizes these functions.

\caption{\label{tab:trigonometric-tb-pdf}cardinalR trigonometric data generation functions}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{8cm}}
\toprule
Function & Explanation\\
\midrule
gen\_crescent & Crescent pattern.\\
gen\_curvycylinder & Curvy cylinder.\\
gen\_sphericalspiral & Spherical spiral.\\
gen\_helicalspiral & Helical spiral.\\
gen\_conicspiral & Conic spiral.\\
\addlinespace
gen\_nonlinear & Nonlinear hyperbola.\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{\label{tab:noise-tb-pdf}cardinalR noise dimensions generation functions}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{8cm}}
\toprule
Function & Explanation\\
\midrule
gen\_noisedims & Gaussian noise dimensions with optional mean and standard deviation.\\
gen\_wavydims1 & Wavy noise dimensions based on a user-specified theta sequence with added jitter.\\
gen\_wavydims2 & Wavy noise dimensions using polynomial transformations of an existing dimension vector.\\
gen\_wavydims3 & Wavy noise dimensions using a combination of polynomial and sine transformations based on the first three dimensions of a dataset.\\
\bottomrule
\end{tabular}
\end{table}

The \texttt{gen\_noisedims(n,\ p,\ m,\ s)} function generates \$p\$ independent Gaussian noise dimensions,

>>>>

@@

A branching structure (Figure \ref{fig:branch-proj}) captures trajectories that diverge or bifurcate from a common origin, similar to processes such as cell differentiation in biology (\citet{trapnell2014}). We introduce a set of data generation functions specifically designed to simulate high-dimensional branching structures with various geometries, numbers of points (\texttt{n}), and number of branches (\texttt{k}). Although these functions can generate multiple branches, they do not produce a formal \emph{multicluster} dataset: the branches form a single connected structure, with multiple visually distinct arms rather than independent clusters.

@@

A cube structure (Figure \ref{fig:cube-proj}) represents uniformly or systematically distributed points within a high-dimensional hypercube, providing a useful framework for assessing how well algorithms preserve uniformity, spacing, and boundary properties in high dimensions. We provide a set of functions to generate high-dimensional cube structures with flexible configurations, including regular grids, uniform random points, and cubes with missing regions or holes. These structures are valuable for testing the ability of algorithms to maintain uniform spacing or to detect gaps in the data.

@@

@@

A polynomial structure (Figure \ref{fig:poly-proj}) generates data points that follow non-linear curvilinear relationships, such as quadratic or cubic trends, in high-dimensional space. These patterns are useful for evaluating how well algorithms capture smooth, non-linear trajectories and curvature in the data. We provide functions for generating quadratic and cubic structures, enabling controlled experiments with different degrees of polynomial complexity.

@@

A pyramid structure (Figure \ref{fig:pyr-proj}) represents data arranged around a central apex and base, useful for exploring how algorithms handle pointed or layered geometries in high-dimensional space. The functions provided allow users to generate pyramids with rectangular, triangular, and star-shaped bases, and sharp or blunted apexes. Additionally, it is possible to create a pyramid with a fractal-like internal structure, enabling the study of non-convex and sparse regions.

@@

@@

An S-curve structure (Figure \ref{fig:scurve-proj}) simulates data that lies along a smooth, non-linear manifold. The functions generate both the standard S-curve shape and, a S-curve variant with structured hole that introduce missing or incomplete region. These variations are useful for evaluating how well algorithms capture non-linear geometry and handle incomplete manifolds in high-dimensional data.

@@

Sphere-shaped structures (Figure \ref{fig:sphere-proj}) are useful for evaluating how dimension reduction and clustering algorithms handle curved, symmetric manifolds in high-dimensional spaces. The functions generate a variety of spherical forms, including simple circles, uniform spheres, grid-based spheres, and complex arrangements like clustered spheres within a larger sphere. The first few coordinates define the main geometric form (circle, cycle, sphere, or hemisphere), while higher-dimensional embeddings are achieved by adding noise dimensions. For the circle and curvy cycle, this occurs when \(p > 2\) and \(p > 3\), respectively, while for spheres and hemispheres it occurs when \(p > 3\) or \(p > 4\).

@@

Trigonometric-based structures provide flexible ways to simulate complex curved patterns and spirals that often arise in real-world high-dimensional data, such as in biological trajectories, or physical systems (Figure \ref{fig:triginometric-proj}). The first few coordinates define the main geometric structure (e.g., crescent, cylinder, spiral, or helix), while for higher-dimensional embeddings, additional noise dimensions are appended: for the crescent shape this occurs when \(p > 2\) (adding \(X_3\) through \(X_p\)), and for all other trigonometric shapes when \(p > 4\) (adding \(X_5\) through \(X_p\)). These structures are particularly valuable for testing how well dimension reduction and clustering algorithms preserve intricate geometric and topological features.

@@

The \texttt{gen\_noisedims(n,\ p,\ m,\ s)} function generates \(p\) independent Gaussian noise dimensions,
